\LARGE \bf Bringing robotics in formal education \\ using the Thymio open source hardware robot

### 2016-04-01 paper_ram.tex
added discussion with e-puck and khepera, + minor issues

Despite this potential, robots are still not widespread in schools as they could be and this for several reasons. Among them we can mention:
\begin{enumerate}
\item Although there are many robotic projects developing educational robots, few reach a sufficient maturity to become distributed and accessible to schools. 
\item A robot performing interesting behaviors is a complex piece of technology and therefore expensive. 
Indeed, for having an educational value and providing an interesting level of interaction, a robot must embed a wide set of sensors and actuators.
Existing platforms with these features cost several hundred Euros.
This prevents most schools, which have a limited budget for equipment, to acquire interesting robots.
\item Introducing robotic tools into teaching activities requires investment in time and training for the teachers.
In Europe, despite a trend in including more and more technological tools into the learning process, teachers are still insufficiently trained and are reluctant to introduce these tools in their teaching activity~\cite{CERI2008}.
To be accepted by teachers, robots must be therefore both accessible with minimal effort and be associated with-well prepared educational material. 
Moreover an infrastructure should allow the exchange of educational material among the teachers.
\item Robot construction, use, and programming is often perceived as a boyish activity in our society~\cite{leonard2009lego,nourbakhsh2009robot}.
% For instance, several authors report that girls have such a perception of the LEGO Minstorms~\cite{leonard2009lego,nourbakhsh2009robot}.
This strongly limits the potential of robots as general-purpose educational tools, especially in schools.
\item Finally, many teachers are reluctant to follow volatile trends, especially if based on purely commercial arguments. Teachers prefer to invest in stable tools, and this is the opposite of todays consumer technology, often very volatile and based on continuos renewal rather than on durability.
\end{enumerate}

@@

There is a very large number of publications presenting educational robots with different prices and features, from very low-cost systems targeting, for instance, education in Africa~\cite{Rubenstein2015,Gyebi2015} to extremely sophisticated humanoids~\cite{Hood2015,Mazzoni2016}.
Among those, only very few are commercially available, limiting their validation by education scientists.
This results in 90\% of publications about validation of educational results made on LEGO Mindstorm~\cite{benitti2012exploring}.
The LEGO Mindstorm EV3\footnote{\url{http://mindstorms.lego.com}} is expensive ($\approx$ 400\,\$) but offers a wide range of possibilities, especially at the mechanical level using LEGO bricks and at the software level with its graphical programming environment. 
At a lower price, LEGO proposes the WeDo system ($\approx$ 80\,\$).
The WeDo is similar to EV3 but only reads one sensor and controls one actuator and must always be connected to a desktop computer.
The BeeBot robot\footnote{\url{http://www.beebot.org.uk}} is a cheap ($\approx$ 90\,\$) autonomous system.
However, it is not a real robot but rather an automaton: There is no perception of the environment, just a good encoder for precise movements and a simple programming interface based on buttons giving directions for the movements.

Among these widely used robots, none is part of an open source hardware project.
The only effort in this direction comes from LEGO, that made public all the schematics of the EV3 in an hardware developer kit, to allow the design of extensions. 
Otherwise, only few educational robots are both used in schools and are issued from an open hardware project:
Scribbler2, produced and sold by Parallax\footnote{\url{http://www.parallax.com}} ($\approx$ 180\,\$), is a 188~mm large robot, designed to run on the ground and equipped with few light sensors, one distance sensor, two ground sensors and few LEDs display.
It runs on standard AA batteries and has a hacker port for interfacing electronic extensions.
It is programmable with a graphic or text code interface.  
The main weakness of Scribbler2 is its performance per price ratio and the limited compatibility with other systems. 
The e-puck~\cite{mondada2009puck} robot is a robot targeting university-level education.
Well equipped with sensors and actuators, modular and compact, it can be programmed with professional environments.
Several simulators allow to run experiments of high complexity. 
It's main weakness is its high price ($\approx$ 870\,\$).
Finch\footnote{\url{http://www.finchrobot.com/}} ($\approx$ 99\,\$) is a very simple robot that has been designed around a wired connection to the computer. 
This connection reduces electronics requirements, such as batteries or wireless communication, and allows to implement the control entirely on the computer.
This results in a very broad set of possible programming languages available, the real force of this robot. 
On the other hand the cable does not allow real autonomy and mobility of the robot.
Finally the mBot\footnote{\url{http://www.makeblock.cc/mbot/}} is a mobile platform based on the Arduino board. 
The very cheap and simple electronics allow to drastically reduce the price of the system ($\approx$ 75\,\$) but also reduces the perception possibilities to a couple of sensors.

@@

For teachers, the motivation to use robotic tools in their classes depends on many factors \cite{chevalier2016}.
Among them, the availability of material and training plays a key role.
The BeeBot mentioned previously, because of its radical simplicity, is one of the tools that needs the least training. 
Moreover its use in learning activities is based on a simple grid on a printed mat on the ground, allowing fast and easy integration in existing lessons.
These are probably two important factors for its success in schools~\cite{2008highfield,2008demichele} and the availability of a rich set of educational material\footnote{\url{https://www.learningplace.com.au/deliver/content.asp?pid=38840} and \\ \url{http://beebots.skola.edu.mt/}}.
However, the total lack of sensors reduces the BeeBot to an automaton, easy to use but limited to very simple activities for young kids, without much potential in teaching computational thinking or robotics concepts.

For more complex robots, both the development of educational material and the training of teachers require a very important effort, requiring a good mix between robotics and educational skills. 
Moreover, educational material changes very much from school to school, as education is very dependent from local educational programs and languages, for instance. 
In the particular case of Switzerland, for instance, every of the 26 states composing the confederation has its own school program and its own language among the four official ones. 
Educational material and the related training have therefore to be developed for each situation, which requires a huge effort.
Even for large companies like LEGO this development is not affordable. 
A solution is to have an active community of users contributing to the development of the material in a distributed manner, adapting the material to the local situation.
LEGO itself is moving in the direction of promoting communities of users \cite{Hienerth2014} with a user-producer interaction close to those of open hardware projects.
The open source community regrouping developers, producers and final users is therefore a very interesting model to address the distributed development and sharing of educational material, and the diffusion of training sessions.
Communities around projects like Arduino are extremely good examples of this type of mechanism \cite{Jamieson2010}.

@@

We designed the Thymio robot along seven main axes: a low price to address a larger number of users, a feature set that suits both genders and multiple ages from young children to adults, a combination of sensors, actuators, and programming features that facilitates learning, a mechanical design that allows creative extensions by the users themselves, a set of ready-to-use behaviors to get a demonstration right out of the box, an accessible programming environment, and an open source  community contributing to the design and disseminating the results.
The result is a miniature (10\,cm) differential-wheeled robot suited for use on a desktop.
The robot is robust enough to be mishandled by children, as it can fall from a table without breaking itself.
It features a translucent white hull and a wide range of sensors and actuators (see \ref{fig:schematique}).
The robot embeds a battery, rechargable by \textsc{usb}, that provides an autonomy of about 2 hours.

@@

Our main concern during the mechatronic design of Thymio was to keep a low production cost. 
Price is a primary factor for the adoption of the robot by schools~\cite{kradolfer2014sociological}.
In our previous experience with designing robots, most of the costs was due to the electronics and the sensors~\cite{thesis_michael}, not to the mechanical parts.
As we wanted Thymio to be an affordable robot, we selected electronic components with great care and explored different design solutions in order to limit the number of expensive components.
In particular, we focused on low-cost sensors that allow rich interaction possibilities, with both the environment and the user.
The resulting Thymio robot possesses a large number of sensors:
Seven horizontal infrared distance sensors, with a range of about 13\,cm.
They enable many interactions with the environment, like following an object or avoiding an obstacle.
Two ground infrared sensors allow to follow a line or to detect the edge of a table.
A three-axis accelerometer measures the orientation of the robot with respect to gravity, and can detect shocks and free falls.
Five capacitive touch buttons organized as an arrow form an intuitive user interface.
Compared to physical buttons, these simplify the plastic hull of the robot and make it more robust.
A remote control receiver provides additional distant buttons.
A microphone can record sounds and detect claps. 
Finally, a thermistor measures temperature.
Most of these sensors cost less then 0.2\,\$, the most expensive being the accelerometer with a cost of about 0.8\,\$, which is an acceptable price given the possibilities it brings to the robot.
The omnipresence of this type of sensors in smartphones and compact cameras has drastically lowered their prices over the last decade.
We also chose low-cost toy motors and control them in speed (max. 13\,cm/s) using a custom-designed electronics.
By measuring the back-electromotive force, speed control is possible without additional encoders.

@@

Thymio finally costs 35\,\$ per unit when produced in a batch of 1\,k units (see \ref{tbl:thymio-price}), with fixed fees of 15\,k\,\$ (13\,k\,\$ for the mold and 2\,k\,\$ for the \textsc{ce} certification).
This allows a selling price of 129\,CHF ($\approx$ 130\,\$).

@@

The square format of the hull allows to use the robot as a base for driving the user's own constructions.
To that end, Thymio is compatible with LEGO bricks:
There are four attachment positions on the top of the robot and two LEGO Technics crosses on each side.
Additionally, the two wheels have attachment points, which permits to use them to actuate elements elsewhere in the structure (see Figure~\ref{fig:example-construction}, third row left) or to lift the robot's own weight (see Figure~\ref{fig:example-construction}, third row right).
Therefore we chose more powerful motors than strictly necessary to move the robot around.
The same fixation points used for LEGO bricks can be used to attach paper structures.
Paper can allow to simply change the body shape and add some body movements, as illustrated in Figure~\ref{fig:example-construction} by the orca, opening and closing its mouth while moving forward or by the bat, moving its wings. 
But paper and cardboard can also radically change the locomotion principle, as illustrated in the second row of Figure~\ref{fig:example-construction} by the zombie, where the robot's wheel activate the legs. 
The paper structure can be also used to interact with the sensors, as illustrated in the second row of Figure~\ref{fig:example-construction} by the bear, who extends his paw in front of the sensors to drive its iceberg.
The same fixation points can be used to attach 3D printed customized parts, as illustrated by the winder shown in the fourth row of Figure~\ref{fig:example-construction}.
Finally one can use paper to create environments, either flat with patterns that can be used in association to the ground sensors (see Figure~\ref{fig:example-construction} bottom) or 3D object, like the trees beside the zombie in Figure~\ref{fig:example-construction}.
Finally, it is also possible to link several Thymio by software, allowing the coordination of complex multi-Thymio robotic structures.

\begin{figure}
\includegraphics[width=.49\columnwidth]{figures/orca}\hfill\includegraphics[width=.49\columnwidth]{figures/bat}
\vskip .5em
\includegraphics[width=.49\columnwidth]{figures/zombie}\hfill\includegraphics[width=.49\columnwidth]{figures/ours}
\vskip .5em
\includegraphics[width=.49\columnwidth]{figures/moving-heli}\hfill\includegraphics[width=.49\columnwidth]{figures/funi}
\vskip .5em
\includegraphics[width=.49\columnwidth]{figures/both_half_winder}\hfill\includegraphics[width=.49\columnwidth]{figures/winder_on_thymio2}
\vskip .5em
\includegraphics[width=\columnwidth]{figures/5senses}
\caption{Examples of extensions of the Thymio basic robot with paper or cardboard body extensions (top four images), using LEGO$^{TM}$ structural extensions (third row), using 3D-printed extensions (fourth row) or using a printed environment (bottom).}
\label{fig:example-construction}
\end{figure}

@@

The choice of developing a robot within an open source hardware project has an impact on the robot design and the way this device is used in the community of users. 
In this section we analyze more in detail the implications of this choice in the context of educational robotics.
We compare these implications with the result of a survey that got 35 answers from people active in various open source hardware projects based on a worldwide call for contributions. 
Among these 35 answers, 11 come from project leaders, 13 from core design team members, 8 from contributors and 3 from enthusiastic users.
54\% of these respondents are between 25 and 35 years old, 26\% between 35 and 50 years old, 9\% are less than 25 years old and 11\% are older than 50.  
67\% either work in an academic environment or are students. 
34 of the 35 answers come from male repondents.
This last point characterizes strongly the hardware developers community.

@@

When asked about their motivation to participate to open source hardware projects, the people participating to the survey give as main motivation a link to their professional activity and to the specific project, followed by a more general motivation by the nature of the project and finally by the engineering practice (see Figure~\ref{fig:motivation}). 
The more abstract goal of improving our society is mentioned only by few respondents.

@@

For the accessories, less technically challenging and stronger linked with creativity and educational value, we promoted techniques that are accessible to ``anyone'' in the broader sense: paper, cardboard, LEGO constructions and 3D printing, as explained in section~\ref{sec:crea} and illustrated in Figure~\ref{fig:example-construction}.
This allows a much broader spectrum of contributor, especially among the teachers.

@@

In some cases, the open source nature of this project was a crucial political argument for the choice of this platform.
In the state of Geneva, for instance, the government decided to move toward open source non-proprietary tools in education.
All primary school- and most secondary school computers have been installed with the Ubuntu distribution of Linux. 
On this operating system the LEGO Mindstorms software tools are not available, and their proprietary nature is not compatible with the philosophy dictated by the government. 
Thymio was therefore extremely welcome, and the state of Geneva organized own training courses for teachers, ported several educational materials into their curriculum and organized a lending system for all schools that cannot afford to buy the robots.
Others states are starting following this model.

>>>>

Despite this potential, robots are still not widespread in schools as they could be and this for several reasons. Among them we can mention:
\begin{enumerate}
\item Although there are many robotic projects developing innovative and interesting educational robots, few reach a sufficient maturity to become distributed and accessible to schools. 
\item A robot performing interesting behaviors is a complex piece of technology and therefore expensive. 
Indeed, for having an educational value and providing an interesting level of interaction, a robot must embed a wide set of sensors and actuators.
Existing platforms with these features cost several hundred Euros.
This prevents most schools, which have a limited budget for equipment, to acquire interesting robots.
\item Introducing robotic tools into teaching activities requires investment in time and training for the teachers.
In Europe, despite a trend in including more and more technological tools into the learning process, teachers are still insufficiently trained and are reluctant to introduce these tools in their teaching activity~\cite{CERI2008}.
To be accepted by teachers, robots must be therefore both accessible with minimal effort and be associated with-well prepared educational material. 
Moreover an infrastructure should allow the exchange of educational material among the teachers.
\item Robot construction, use, and programming is often perceived as a boyish activity in our society~\cite{leonard2009lego,nourbakhsh2009robot}.
% For instance, several authors report that girls have such a perception of the LEGO Minstorms~\cite{leonard2009lego,nourbakhsh2009robot}.
This strongly limits the potential of robots as general-purpose educational tools, especially in schools.
\item Finally, many teachers are reluctant to follow volatile trends, especially if based on purely commercial arguments. Teachers prefer to invest in stable tools, and this is the opposite of todays consumer technology, often very volatile and based on continuos renewal rather than on durability.
\end{enumerate}

@@

There is a very large number of publications presenting educational robots with different prices and features, from very low-cost systems targeting, for instance, education in Africa~\cite{Rubenstein2015,Gyebi2015} to extremely sophisticated humanoids~\cite{Hood2015,Mazzoni2016}.
Among those, only very few are commercially available, limiting their validation by education scientists.
This results in 90\% of publications about validation of educational results made on LEGO\textsuperscript{\textregistered} Mindstorms\textsuperscript{\textregistered}~\cite{benitti2012exploring}.
The LEGO\textsuperscript{\textregistered} Mindstorms\textsuperscript{\textregistered} EV3\footnote{\url{http://mindstorms.lego.com}} is expensive ($\approx$ 400\,\$) but offers a wide range of possibilities, especially at the mechanical level using LEGO\textsuperscript{\textregistered} bricks and at the software level with its graphical programming environment. 
At a lower price, LEGO\textsuperscript{\textregistered} proposes the WeDo\textsuperscript{\textregistered} system ($\approx$ 80\,\$).
The WeDo\textsuperscript{\textregistered} is similar to EV3 but only reads one sensor and controls one actuator and must always be connected to a desktop computer.
The BeeBot robot\footnote{\url{http://www.beebot.org.uk}} is a cheap ($\approx$ 90\,\$) autonomous system.
However, it is not a real robot but rather an automaton: There is no perception of the environment, just a good encoder for precise movements and a simple programming interface based on buttons giving directions for the movements.

Among these widely used robots, none is part of an open source hardware project.
The only effort in this direction comes from LEGO\textsuperscript{\textregistered}, that made public all the schematics of the EV3 in an hardware developer kit, to allow the design of extensions. 
Otherwise, only few educational robots are both used in schools and are issued from an open hardware project:
Scribbler2, produced and sold by Parallax\footnote{\url{http://www.parallax.com}} ($\approx$ 180\,\$), is a 188~mm large robot, designed to run on the ground and equipped with few light sensors, one distance sensor, two ground sensors and few LEDs display.
It runs on standard AA batteries and has a hacker port for interfacing electronic extensions.
It is programmable with a graphic or text code interface.  
The main weakness of Scribbler2 is its performance per price ratio and the limited compatibility with other systems. 
Moreover there seems to be no real community around its development.
The e-puck~\cite{mondada2009puck} robot is a robot targeting university-level education.
Well equipped with sensors and actuators, modular and compact, it can be programmed with professional environments.
Several simulators allow to run experiments of high complexity. 
There is an active community of users and developers around it.
It's main weakness is its high price ($\approx$ 870\,\$).
Finch\footnote{\url{http://www.finchrobot.com/}} ($\approx$ 99\,\$) is a very simple robot that has been designed around a wired connection to the computer. 
This connection reduces electronics requirements, such as batteries or wireless communication, and allows to implement the control entirely on the computer.
This results in a very broad set of possible programming languages available, the real force of this robot. 
On the other hand the cable does not allow real autonomy and mobility of the robot.
Finally the mBot\footnote{\url{http://www.makeblock.cc/mbot/}} is a mobile platform based on the Arduino board. 
The very cheap and simple electronics allow to drastically reduce the price of the system ($\approx$ 75\,\$) but also reduces the perception possibilities to a couple of sensors.

@@

Among the main target users there are teachers.
For teachers, the motivation to use robotic tools in their classes depends on many factors \cite{chevalier2016}.
Among them, the availability of material and training plays a key role.
The BeeBot mentioned previously, because of its radical simplicity, is one of the tools that needs the least training. 
Moreover its use in learning activities is based on a simple grid on a printed mat on the ground, allowing fast and easy integration in existing lessons.
These are probably two important factors for its success in schools~\cite{2008highfield,2008demichele} and the availability of a rich set of educational material\footnote{\url{https://www.learningplace.com.au/deliver/content.asp?pid=38840} and \\ \url{http://beebots.skola.edu.mt/}}.
However, the total lack of sensors reduces the BeeBot to an automaton, easy to use but limited to very simple activities for young kids, without much potential in teaching computational thinking or robotics concepts.

For more complex robots, both the development of educational material and the training of teachers require a very important effort, needing a good mix between robotics and educational skills. 
Moreover, educational material changes very much from school to school, as education is very dependent from local educational programs and languages, for instance. 
In the particular case of Switzerland, for instance, every of the 26 states composing the confederation has its own school program and its own language among the four national ones. 
Educational material and the related training have therefore to be adapted to each situation, which requires a huge effort.
Even for large companies like LEGO\textsuperscript{\textregistered} this development is not affordable. 
A solution is to have an active community of users contributing to the development of the material in a distributed manner, adapting the material to the local situation.
LEGO\textsuperscript{\textregistered} itself is moving in the direction of promoting communities of users \cite{Hienerth2014} with a user-producer interaction close to those of open hardware projects.
The open source community regrouping developers, producers and final users is therefore a very interesting model to address the distributed development and sharing of educational material, and the diffusion of training sessions.
This model is an extension to education of what already practiced in projects like Arduino~\cite{Jamieson2010}.

@@

We designed the Thymio robot along seven main axes: a low price to address a larger number of users, a feature set that suits both genders and multiple ages from young children to adults, a combination of sensors, actuators, and programming features that facilitates learning, a mechanical design that allows creative extensions by the users themselves, a set of ready-to-use behaviors to get a demonstration right out of the box, an accessible programming environment, and an open source  community contributing to the design and disseminating the results.
The result is a miniature (10\,cm in diameter, see Figure~\ref{fig:schematique}) differential-wheeled robot suited for use on a desktop.
The robot is robust enough to be mishandled by children, as it can fall from a table without breaking itself.
It features a translucent white hull and a wide range of sensors and actuators (see Figure~\ref{fig:schematique} bottom).
The robot embeds a battery, rechargable by \textsc{usb}, that provides an autonomy of about 2 hours.

@@

Our main concern during the mechatronic design of Thymio was to keep a low production cost. 
Price is a key factor for the adoption of the robot by schools~\cite{kradolfer2014sociological}.
In our previous experience with designing robots, most of the costs was due to the electronics and the sensors~\cite{thesis_michael}, not to the mechanical parts.
As we wanted Thymio to be an affordable robot, we selected electronic components with great care and explored different design solutions in order to limit the number of expensive components.
In particular, we focused on low-cost sensors that allow rich interaction possibilities, with both the environment and the user.
The resulting Thymio robot possesses a large number of sensors:
Seven horizontal infrared distance sensors, with a range of about 13\,cm.
They enable many interactions with the environment, like following an object or avoiding an obstacle.
Two ground infrared sensors allow to follow a line or to detect the edge of a table.
A three-axis accelerometer measures the orientation of the robot with respect to gravity, and can detect shocks and free falls.
Five capacitive touch buttons organized as an arrow form an intuitive user interface.
Compared to physical buttons, these simplify the plastic hull of the robot and make it more robust.
A remote control receiver provides additional distant buttons.
A microphone can record sounds and detect claps. 
Finally, a thermistor measures temperature.
Most of these sensors cost less then 0.2\,\$, the most expensive being the accelerometer with a cost of about 0.8\,\$, which is an acceptable price given the possibilities it brings to the robot.
The omnipresence of this type of sensors in smartphones and compact cameras has drastically lowered their prices over the last decade.
We also chose low-cost toy motors and control them in speed (max. 13\,cm/s) using a custom-designed electronics.
By measuring the back-electromotive force, speed control is possible without additional encoders.

@@

Thymio finally costs 35\,\$ per unit when produced in a batch of 1\,k units (see Table~\ref{tbl:thymio-price}), with fixed fees of 15\,k\,\$ (13\,k\,\$ for the mold and 2\,k\,\$ for the \textsc{ce} certification).
This allows a selling price of 129\,CHF ($\approx$ 130\,\$).

@@

The square format of the hull allows to use the robot as a base for driving the user's own constructions.
To that end, Thymio is compatible with LEGO\textsuperscript{\textregistered} bricks:
There are four attachment positions on the top of the robot and two LEGO\textsuperscript{\textregistered} Technics crosses on each side.
Additionally, the two wheels have attachment points, which permits to use them to actuate elements elsewhere in the structure (see Figure~\ref{fig:example-construction}, third row left) or to lift the robot's own weight (see Figure~\ref{fig:example-construction}, third row right).
Therefore we chose more powerful motors than strictly necessary to move the robot around.
The same fixation points used for LEGO\textsuperscript{\textregistered} bricks can be used to attach paper structures.
Paper can allow to simply change the body shape and add some body movements, as illustrated in Figure~\ref{fig:example-construction} by the orca, opening and closing its mouth while moving forward or by the bat, moving its wings. 
But paper and cardboard can also radically change the locomotion principle, as illustrated in the second row of Figure~\ref{fig:example-construction} by the zombie, where the robot's wheel activate the legs. 
The paper structure can be also used to interact with the sensors, as illustrated in the second row of Figure~\ref{fig:example-construction} by the bear, who extends his paw in front of the sensors to drive its iceberg.
The same fixation points can be used to attach 3D printed customized parts, as illustrated by the winder shown in the fourth row of Figure~\ref{fig:example-construction}.
Finally one can use paper to create environments, either flat with patterns that can be used in association to the ground sensors (see Figure~\ref{fig:example-construction} bottom) or 3D object, like the trees beside the zombie in Figure~\ref{fig:example-construction}.
Finally, it is also possible to link several Thymio by software, allowing the coordination of complex multi-Thymio robotic structures.

\begin{figure}
\includegraphics[width=.49\columnwidth]{figures/orca}\hfill\includegraphics[width=.49\columnwidth]{figures/bat}
\vskip .5em
\includegraphics[width=.49\columnwidth]{figures/zombie}\hfill\includegraphics[width=.49\columnwidth]{figures/ours}
\vskip .5em
\includegraphics[width=.49\columnwidth]{figures/moving-heli}\hfill\includegraphics[width=.49\columnwidth]{figures/funi}
\vskip .5em
\includegraphics[width=.49\columnwidth]{figures/both_half_winder}\hfill\includegraphics[width=.49\columnwidth]{figures/winder_on_thymio2}
\vskip .5em
\includegraphics[width=\columnwidth]{figures/5senses}
\caption{Examples of extensions of the Thymio basic robot with paper or cardboard body extensions (top four images), using LEGO\textsuperscript{\textregistered} structural extensions (third row), using 3D-printed extensions (fourth row) or using a printed environment (bottom).}
\label{fig:example-construction}
\end{figure}

@@

The choice of developing a robot within an open source hardware project has an impact on the robot design and the way this device is used in the community of users. 
In this section we analyze more in detail the implications of this choice in the context of educational robotics.
We compare these implications with the result of a survey that got 35 answers from people active in various open source hardware projects based on a worldwide call for contributions. 
Among these 35 answers, 11 come from project leaders, 13 from core design team members, 8 from contributors and 3 from enthusiastic users.
54\% of these respondents are between 25 and 35 years old, 26\% between 35 and 50 years old, 9\% are less than 25 years old and 11\% are older than 50.  
67\% either work in an academic environment or are students. 
34 of the 35 answers come from male respondents.

@@

We can distinguish two levels of motivation, the institutional and the personal one.
The institution initiating a project is mainly looking for recognition and/or money. 
Our group has a good experience in disseminating robotic hardware with the Khepera~\cite{MonFraIen93} and e-puck~\cite{mondada2009puck} robots.
Khepera was disseminated with a proprietary strategy, e-puck with an open source hardware one.
Both were targeting similar users and have been sold in similar quantities.
What we can observe after more than 10 years is that Khepera generated many royalties for the university, but less relative academic visibility than e-puck (see Figure~\ref{fig:khepuck}) .
The e-puck robot, being open hardware and with an image better linked to the university, generated no income for the university but much more visibility in respect to his diffusion.
Khepera was more linked with the name of the company producing it, the e-puck robot was nearly not linked with the producer when mentioned in the literature.
In the case of Thymio, developed in a research project, the institutional motivation was to get visibility more than money. 
Therefore, based on the past experience, the open source hardware strategy seemed more adapted.

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/others_robots}
\caption{Comparison between mentions in the literature, mention of the manufacturer and citation of the reference paper for two robots disseminated by our lab in the past. It appears that the Khepera robot, distributed with a proprietary strategy, was more linked to the company producing it than the original designers. This is the opposite for the e-puck robot, distributed with an open source license.}
\label{fig:khepuck}
\end{figure}

The personal motivation to participate to such a project is very different than the institutional one. 
When asked about their personal motivation, the people participating to the survey give as main motivation a link to their professional activity and to the specific project, followed by a more general motivation by the nature of the project and finally by the engineering practice (see Figure~\ref{fig:motivation}). 
The more abstract goal of improving our society is mentioned only by few respondents.

@@

For the accessories, less technically challenging and stronger linked with creativity and educational value, we promoted techniques that are accessible to ``anyone'' in the broader sense: paper, cardboard, LEGO\textsuperscript{\textregistered} constructions and 3D printing, as explained in section~\ref{sec:crea} and illustrated in Figure~\ref{fig:example-construction}.
This allows a much broader spectrum of contributor, especially among the teachers.

@@

In some cases, the open source nature of this project was a crucial political argument for the choice of this platform.
In the state of Geneva, for instance, the government decided to move toward open source non-proprietary tools in education.
All primary school- and most secondary school computers have been installed with the Ubuntu distribution of Linux. 
On this operating system the LEGO\textsuperscript{\textregistered} Mindstorms\textsuperscript{\textregistered} software tools are not available, and their proprietary nature is not compatible with the philosophy dictated by the government. 
Thymio was therefore extremely welcome, and the state of Geneva organized own training courses for teachers, ported several educational materials into their curriculum and organized a lending system for all schools that cannot afford to buy the robots.
Others states are starting following this model.
